---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### First

We start by charge the libaries that we will use

```{r}
library(dplyr)
library(ggplot2)
```
### second

We loading the file and turning it into a df table to facilitate its processing and see the first rows.

```{r}
activity<- read.csv("/Users/johnsprockel/Documents/GitHub/ProgrammingAssignment2/activity.csv")
activity<- tbl_df(activity)
head(activity)
```
### Third

We review the summary of the file finding that:

-Its length is 17568 data

-has 2304 NA's

Now we proceed to eliminate NA's 

```{r}
summary (activity)
activity<- activity[!is.na(activity$steps),]
```

-the minimum value of steps is 0 and the maximum is 806, its mean is 37 and median is 0

-the minimum value of interval is 0 and the maximum is 2355, its mean and median are 1177.5.

### Fourth

We proceed to change the date variable to said format to obtain the tidy data.

```{r}
activity$date<- as.Date(as.character(activity$date))
summary(activity)
```

## What is mean total number of steps taken per day?

Initially the table is grouped by date, then we generate a new table with the sum, mean and median of steps for each day.

```{r}
activity<- group_by(activity, date)
x<- summarise(activity, sum= sum(steps), mean = mean (steps), median = median(steps))
x
```

The total steps of each day are plotted.

```{r, fig.height=4}
g<-ggplot(x, aes(date, sum)) 
g+ geom_bar(stat="identity", fill = "red") +
  ggtitle("Total Number of Steps Taken Each Day") +
  xlab("Date") + ylab("Sum Steps")
```

The mean of total steps of each day are plotted.

```{r, fig.height=4}
g<-ggplot(x, aes(date, mean)) 
g+ geom_bar(stat="identity", fill = "steelblue") +
  ggtitle("Mean of Total Number of Steps Taken Each Day") +
  xlab("Date") + ylab("Mean Steps")
```

The mean of total steps of each day are plotted.

```{r, fig.height=4}
g<-ggplot(x, aes(date, median)) 
g+ geom_bar(stat="identity", fill = "steelblue") +
  ggtitle("Median of Total Number of Steps Taken Each Day") +
  xlab("Date") + ylab("Median Steps")
```

Each day, the median was 0.

## What is the average daily activity pattern?

This is a time series plot of the average number of steps taken.

```{r}
g<-ggplot(x, aes(date, mean)) 
g+ geom_line(color = "steelblue") +
  ggtitle("Mean of Total Number of Steps Taken Each Day") +
  xlab("Date") + ylab("Mean Steps")+
  theme_bw(base_family = "Times")
```

## The 5-minute interval that, on average, contains the maximum number of steps

Create an data table with steps and interval and grouped by interval, then calculate the mean for each interval.

```{r}
con<- cbind(activity$steps, activity$interval)
con<- as.data.frame(con)
con<- group_by(con, con[,2])

y<- summarise(con, mean = mean (V1))
colnames(y)<- c("steps", "interval")
y
```

Now plot the average by interval

```{r}
g<-ggplot(y, aes(interval, steps)) 
g+ geom_line(color = "green") +
  ggtitle("Mean of Total Number of Steps Taken by Interval") +
  xlab("Interval") + ylab("Mean Steps")+
  theme_bw(base_family = "Times")
```

## Imputing missing values

We load the data again

```{r}
activity<- read.csv("/Users/johnsprockel/Documents/GitHub/ProgrammingAssignment2/activity.csv")
activity<- tbl_df(activity)

activity$date<- as.Date(as.character(activity$date))
str(activity$steps)
```

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

There are 2304 NA's

2. I decide to filling in all of the missing values in the data set by use the mean for that  5-minute interval.

3. Next, I create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
tt<- activity %>%
  group_by(interval) %>%
	summarise(mean_steps = mean(steps, na.rm = TRUE))

for (i in 1:length(activity$steps)){
  if(is.na(activity$steps[i])){
    activity$steps[i]<- tt$mean_steps[match(activity$interval[i], tt$interval)]
  }
}

```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
activity<- group_by(activity, date)

x<- summarise(activity, sum= sum(steps), mean = mean (steps), median = median(steps))
x
```

The total steps of each day are plotted.

```{r, fig.height=4}
g<-ggplot(x, aes(date, sum)) 
g+ geom_bar(stat="identity", fill = "red") +
  ggtitle("Total Number of Steps Taken Each Day") +
  xlab("Date") + ylab("Sum Steps")
```

The mean of total steps of each day are plotted.

```{r, fig.height=4}
g<-ggplot(x, aes(date, mean)) 
g+ geom_bar(stat="identity", fill = "steelblue") +
  ggtitle("Mean of Total Number of Steps Taken Each Day") +
  xlab("Date") + ylab("Mean Steps")
```


## Are there differences in activity patterns between weekdays and weekends?

Firts I create a new column with a factor variable in the dataset with the days of the week, then clasified in two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
activity<- read.csv("/Users/johnsprockel/Documents/GitHub/ProgrammingAssignment2/activity.csv")
activity<- tbl_df(activity)
activity<- activity[!is.na(activity$steps),]
activity$date<- as.Date(as.character(activity$date))

activity<- mutate(activity, day= weekdays(activity$date))
activity$day<- gsub("Saturday", "Weekend", activity$day)
activity$day<- gsub("Sunday", "Weekend", activity$day)
activity$day<- gsub("Friday", "Weekday", activity$day)
activity$day<- gsub("Monday", "Weekday", activity$day)
activity$day<- gsub("Thursday", "Weekday", activity$day)
activity$day<- gsub("Tuesday", "Weekday", activity$day)
activity$day<- gsub("Wednesday", "Weekday", activity$day)
```

Next I create two tables with the average number of steps taken in weekdays and weekends, by interval, that finally bind in a final database (db)

```{r}

WD<- cbind(unique (activity$interval), "Weekday", 0)
colnames(WD)<- c("interval", "day", "steps")
WD<- as.data.frame(WD)
WD[,1] <- as.numeric(WD[,1])
WD[,3] <- as.numeric(WD[,3])

WE<- cbind(unique (activity$interval), "Weekend", 0)
colnames(WE)<- c("interval", "day", "steps")
WE<- as.data.frame(WE)
WE[,1] <- as.numeric(WE[,1])
WE[,3] <- as.numeric(WE[,3])

for (i in 1:288){
  WD$steps[i]<- mean(activity$steps[activity$interval[]==WD$interval[i]] & activity$day[]== "Weekday")
  WE$steps[i]<- mean(activity$steps[activity$interval[]==WE$interval[i]] & activity$day[]== "Weekend")
}
head(WD)

head(WE)

db<- rbind(WD, WE)
str(db)
```

We proceed to graph the relationship of the different steps by interval in weekdays and weekends. 

```{r}
g<-ggplot(db, aes(interval, steps)) 
g+ geom_line(color = "steelblue") +
  ggtitle("Mean in Total Number of Steps Taken by Weekday") +
  xlab("Interval") + ylab("Sum Steps")+
  facet_grid(day~.)
```